---
title: "What is the light cytosol?"
author:
  - name: "Tom Smith"
    affiliation: "Cambridge Centre for Proteomics"
date: "`r format(Sys.time(), '%d %B, %Y')`"
abstract: | 
  Here, we consider the RNAs in the 'light cytosol' to try and understand what 
  this profile represents
output:
  pdf_document:
  html_notebook: default
geometry: margin=1in
fontsize: 11pt
---


```{r}
library(tidyverse)
library(camprotR)
library(pRoloc)
library(pRolocdata)
library(pRolocExt)

```
Read in the proportions and dLoRNA quantification

```{r}
proportions <- readRDS('./out/proportions_single_table.rds')
proportions_per_rep <- readRDS('./out/proportions_per_rep_single_table_flt.rds')
```

```{r}
dlorna <- readRDS('../2_clustering/out/all_datasets_with_novel.rds')

```

Annotate the proportions with features from quantification object.
```{r}
proportions_annotated <- proportions %>% 
  names() %>%
  lapply(function(level){
    proportion_annotated <- proportions[[level]] %>%
      rowwise() %>%
      mutate(Nuc=sum(Nucleus, Nucleolus, na.rm=TRUE)) %>%
      dplyr::select(id, condition, Cytosol, Nuc, Membrane, Cytosol_Light)
    
    if(level=='transcript'){
      proportion_annotated <- proportion_annotated %>%
        merge(fData(dlorna[[level]]$DMSO$corrected)[
          , c('ensembl_gene_id', 'gene_biotype', 'transcript_biotype',
              'external_gene_name', 'chromosome_name', 'transcript_length')],
        by.x='id', by.y='row.names')
    } else{
      proportion_annotated <- proportion_annotated %>%
        merge(fData(dlorna[[level]]$DMSO$corrected)[
          , c('gene_biotype', 'external_gene_name', 'chromosome_name')],
        by.x='id', by.y='row.names')
      
    }
    
    return(proportion_annotated)
    
  })

names(proportions_annotated) <- names(proportions)

proportions %>% lapply(dim)
proportions_annotated %>% lapply(dim)
```

```{r, fig.height=10, fig.width=10}
mean_proportions_to_plot <- proportions_annotated %>%
  names() %>%
  lapply(function(level){
    if(level=='transcript'){
      biotype_col='transcript_biotype'
    } else {biotype_col='gene_biotype'}
    
    
    proportions_annotated[[level]] %>%
        filter(!!sym(biotype_col) %in% c('protein_coding', 'lncRNA',
                                         'nonsense_mediated_decay', 'processed_transcript',
                                         'retained_intron')) %>%
      mutate(biotype=!!sym(biotype_col)) %>%
      pivot_longer(cols=c(Cytosol, Cytosol_Light, Membrane, Nuc),
                   names_to='localisation',
                   values_to='proportion')
  })
names(mean_proportions_to_plot) <- names(proportions_annotated)  

```

Read in the ribosome profiling data from Teemu et al (2015).
```{r}

ribosome_profiling <- readRDS('../../../../../1_external/2_ribosome_profiling/out/Teemu_et_al_2015_ribosome_profiling.rds') %>%
  dplyr::select(ribo_density)

to_plot <- proportions_annotated$gene %>%
  filter(condition=='DMSO') %>%
  dplyr::select(id, Cytosol_Light) %>%
  merge(ribosome_profiling, by.x='id', by.y='row.names')
  
to_plot %>%
  ggplot(aes(Cytosol_Light, log2(ribo_density))) +
  geom_point(size=0.25, alpha=0.25) +
  theme_camprot() +
  geom_smooth()

p <- to_plot %>%
  ggplot(aes(Hmisc::cut2(Cytosol_Light, g=10),
             log2(ribo_density))) +
  geom_boxplot(notch=TRUE) +
  theme_camprot(border=FALSE, base_family='sans', base_size=15) +
  ylab('Ribosome density (log2)') +
  xlab('\n\n\n') +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())

print(p)

range <- log2(to_plot$ribo_density)
range <- range[is.finite(range)]

range_range <- max(range)-min(range)
tri_min <- max(range) - (range_range*1.15)
tri_max <- max(range) - (range_range*1.1)
  
p2 <- p +
  annotate(geom='polygon', x=c(1,10,10,1), y=c(tri_min, tri_min, tri_max, tri_min), fill='grey40')  +
  annotate(geom='text', x=5.5, y=(max(range) - (range_range*1.2)), label='Cytosol light proportion', size=5, vjust=0.5) +
  coord_cartesian(ylim=c(min(range), max(range)), clip = 'off')

print(p2)

ggsave('../../../../../5_manuscript_figures/Figure_5/ribo_density_length/Ribosome_density.png', p2, width=3.5, height=3.5)
ggsave('../../../../../5_manuscript_figures/Figure_5/ribo_density_length/Ribosome_density.pdf', p2, width=3.5, height=3.5)
```
OK, so there's a clear relationship between cytosol light proportion and the ribsome density, with ~8-fold difference between the bottom and top 10% by cytosol light proportion.

Let's also look at the relationship between transcript length and cytosol light proportion.
```{r}
p <- proportions_annotated$transcript %>%
  filter(condition=='DMSO') %>%
  ggplot(aes(Hmisc::cut2(Cytosol_Light, g=10),
             log10(transcript_length))) +
  geom_boxplot(notch=TRUE) +
  theme_camprot(border=FALSE, base_family='sans', base_size=15) +
  ylab('Transcript length (nt;log10)') +
  xlab('\n\n\n') +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())

print(p)

range <- proportions_annotated$transcript %>%
  filter(condition=='DMSO') %>%
  pull(transcript_length) %>%
  log10()
range <- range[is.finite(range)]

range_range <- max(range)-min(range)
tri_min <- max(range) - (range_range*1.15)
tri_max <- max(range) - (range_range*1.1)
  
p2 <- p +
  annotate(geom='polygon', x=c(1,10,10,1), y=c(tri_min, tri_min, tri_max, tri_min), fill='grey40')  +
  annotate(geom='text', x=5.5, y=(max(range) - (range_range*1.2)), label='Cytosol light proportion', size=5, vjust=0.5) +
  coord_cartesian(ylim=c(min(range), max(range)), clip = 'off')

print(p2)

ggsave('../../../../../5_manuscript_figures/Figure_5/ribo_density_length/length.png', p2, width=3.5, height=3.5)
ggsave('../../../../../5_manuscript_figures/Figure_5/ribo_density_length/length.pdf', p2, width=3.5, height=3.5)

```



```{r}
deseq_results <- readRDS('../../1_Differential_abundance/results/DESeq2_results_no_lf_compression.rds')
```

```{r, fig.height=10, fig.width=10}
to_plot_vs_exp_diff <- mean_proportions_to_plot %>%
  names() %>%
  lapply(function(level){
    x_key <- ifelse(level=='transcript', 'ensembl_gene_id', 'id')

    mean_proportions_to_plot[[level]] %>%
      #dplyr::select(id, ensembl_gene_id, biotype, condition, localisation, proportion) %>%
      pivot_wider(names_from=condition, values_from=proportion) %>%
      merge(deseq_results, by.x=x_key, by.y='row.names') %>%
      mutate(diff=Tg-DMSO)
  })

names(to_plot_vs_exp_diff) <- names(mean_proportions_to_plot)

to_plot_vs_exp_diff %>%
  names() %>%
  lapply(function(level){
    to_plot_vs_exp_diff[[level]] %>%
      filter(baseMean>2^10, biotype %in% c('lncRNA', 'protein_coding')) %>%
      ggplot(aes(log2FoldChange, diff)) +
      geom_point(aes(size=biotype, alpha=biotype)) +
      theme_camprot() +
      scale_size_manual(values=c(0.5, 0.1), guide=FALSE) +
      scale_alpha_manual(values=c(0.5, 0.1), guide=FALSE) +
      facet_grid(biotype~localisation) +
      geom_vline(xintercept=0, linetype=2, colour='grey') +
      geom_hline(yintercept=0, linetype=2, colour='grey') +
      geom_smooth() +
      coord_cartesian(xlim=c(-0.5, 1)) +
      xlab('Change in RNA abundance (log2)') +
      ylab('Change in localisation')
  })
```
OK, so it looks like those protein-coding genes/transcripts that start in the membrane and move towards the Cytosol_Light might have slight increase in abundance overall. That would be unexpected. Let's look at that more directly
```{r}
mean_proportions_to_plot %>% names() %>%
  lapply(function(level){
  
  membrane_rnas <- mean_proportions_to_plot[[level]] %>%
    filter(biotype=='protein_coding', condition=='DMSO', localisation=='Membrane', proportion>=0.5) %>%
    pull(id)
  
  to_plot_vs_exp_diff[[level]] %>%
    filter(localisation=='Cytosol_Light', id %in% membrane_rnas) %>%
    filter(baseMean>2^10) %>%
    ggplot(aes(diff, log2FoldChange)) +
    geom_point() +
    theme_camprot() +
    geom_vline(xintercept=0, linetype=2, colour='grey') +
    geom_hline(yintercept=0, linetype=2, colour='grey') +
    geom_smooth() +
    ggtitle(level) +
    #coord_cartesian(ylim=c(-0.5, 1)) +
    ylab('Change in RNA abundance (log2)') +
    xlab('Change in Cytosol Light proportion')
})
```
Hmmm, so no relationship whatsoever between the amount of movement towards the cytosol light and the change in RNA abundance. So the overall higher abundance for membrane-localised RNAs is independent of the relocalisation. An interesting observation nonetheless.


```{r}


Hubstenberger_et_al <- readRDS('../../../../../1_external/3_granules/Hubstenberger_et_al_2017/hub_p_body_rna.rds')
Hubstenberger_et_al_enr <- Hubstenberger_et_al %>% filter(enr>3, fdr<0.01)
print(nrow(Hubstenberger_et_al_enr))
```

```{r}
colours <- readRDS('../../../../../6_shiny_app/out/shiny_colours.rds')$RNA[1:6]

obj <- dlorna$gene$DMSO$corrected
fData(obj)$markers_inc_hub <- fData(obj)$markers
fData(obj)$markers_inc_hub[
  ((getMarkers(obj, 'markers')=='unknown') &
     (rownames(obj) %in% Hubstenberger_et_al_enr$ensembl_gene_id))] <- 'Hubstenberger et al P-body'
fData(obj)$markers_inc_hub <- factor(fData(obj)$markers_inc_hub,
                                       levels=c(getMarkerClasses(obj, 'markers'),
                                                c('Hubstenberger et al P-body', 'unknown')))


p <- mrkConsProfiles(obj, fcol='markers_inc_hub') %>%
  data.frame() %>%
  tibble::rownames_to_column('markers') %>%
  pivot_longer(cols=-markers) %>%
  mutate(markers=factor(remove_x(markers), levels=levels(fData(obj)$markers_inc_hub))) %>%
  mutate(markers=recode(markers, 'Cytosol_Light'='Granule')) %>%
  merge(pData(obj), by.x='name', by.y='sample') %>%
  ggplot(aes(fraction, value, colour=markers, group=interaction(markers, replicate))) +
  geom_line(size=1) +
  theme_camprot(border=FALSE, base_size=15, base_family='sans' ) +
  facet_wrap(~(paste0('Rep. ', replicate)), nrow=1) +
  scale_colour_manual(values=c(colours, 'black'), name='') +
  theme(strip.background=element_blank()) +
  xlab('Fraction') +
  ylab('Mean abundance\n(sum norm.)')
      
print(p)
ggsave('../../../../../5_manuscript_figures/Figure_5/external/Hubstenberger.png', width=9, height=3.5)
ggsave('../../../../../5_manuscript_figures/Figure_5/external/Hubstenberger.pdf', width=9, height=3.5)
```




```{r}

for(condition in names(dlorna$gene)){
  for(correction in names(dlorna$gene[[condition]])){
  .data <- dlorna$gene[[condition]][[correction]]

    title <- paste(condition, correction, sep=', ')
    
    fData(.data)$markers_inc_hub <- fData(.data)$markers
    fData(.data)$markers_inc_hub[
      ((getMarkers(.data, 'markers')=='unknown') &
         (rownames(.data) %in% Hubstenberger_et_al_enr$ensembl_gene_id))] <- 'Hubstenberger P-body'
    fData(.data)$markers_inc_hub <- factor(fData(.data)$markers_inc_hub,
                                           levels=c(getMarkerClasses(.data, 'markers'),
                                                    c('Hubstenberger P-body', 'unknown')))
    
    .pca <- plot2D(.data, fcol='markers_inc_hub', col=c(colours, 'black'), main=title)
    addLegend(.data, where='bottomleft', fcol='markers_inc_hub', col=c(colours, 'black'))
    
    p <- plot_marker_profiles(.data[fData(.data)$markers_inc_hub!='unknown',], 'markers_inc_hub',
                         group_by='replicate', alpha=c(rep(0.25, 6), 0.05), facet_by='markers') +
      scale_colour_manual(values=c(colours, 'black')) +
      scale_x_discrete(labels=pData(.data)$fraction) +
      ggtitle(title)
    
    
    p2 <- plot_marker_profiles(.data[fData(.data)$markers_inc_hub!='unknown',], 'markers_inc_hub',
                         group_by='replicate', alpha=c(rep(0.25, 6), 0.05)) +
      scale_colour_manual(values=c(colours, 'black')) +
      scale_x_discrete(labels=pData(.data)$fraction) +
      ggtitle(title)
    
    p3 <- plot_marker_profiles(.data[fData(.data)$markers_inc_hub!='unknown',], 'markers_inc_hub',
                         group_by='replicate', plot_all=FALSE) +
      scale_colour_manual(values=c(colours, 'black')) +
      scale_x_discrete(labels=pData(.data)$fraction) +
      ggtitle(title)

    print(p)
    print(p2)
    print(p3)
  }
  
}

```

Comparison with Khong et al: https://pubmed.ncbi.nlm.nih.gov/29129640/
```{r}
khong_et_al <- readxl::read_excel('../../../../../1_external/3_granules/Khong_et_al_2017/1-s2.0-S1097276517307906-mmc2.xlsx', sheet=1)

p <- mean_proportions_to_plot$gene %>% filter(condition=='Tg', localisation=='Cytosol_Light') %>%
  merge(khong_et_al, by.x='id', by.y='test_id') %>%
  mutate(Localization=factor(Localization, levels=c('enriched', 'Neither', 'depleted'))) %>%
  ggplot(aes(proportion, x=log2(`Fold change`))) +
  geom_point(aes(colour=Localization, alpha=Localization), size=0.5, alpha=0.1) +
  geom_smooth() +
  theme_camprot(border=FALSE, base_family='sans') +
  xlim(-5, NA) +
  scale_colour_manual(values=c(get_cat_palette(1), 'grey', get_cat_palette(2)[2]), name='Khong et al (2017)',
                      labels=c('Enriched', 'Neither', 'Depleted')) +
  scale_alpha_manual(values=c(1, 0.1, 1), name='Khong et al',
                      labels=c('Enriched', 'Neither', 'Depleted')) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size=3))) +
  ylab('LoRNA cytosol light\nproportion (UPR)') +
  xlab('Khong et al (2017)\nSG/Total (log2)') +
  ylim(0,1)

print(p)

ggsave('../../../../../5_manuscript_figures/Figure_5/external/Khong.png', width=4, height=4)
ggsave('../../../../../5_manuscript_figures/Figure_5/external/Khong.pdf', width=4, height=4)
```



Any relationship between relocalisation to membrane and change in RNA abundance?
```{r}
RNA_abundance_change <- readRDS('../../1_Differential_abundance/results/DESeq2_results.rds')
```

```{r}

proportions$gene %>%
  dplyr::select(id, condition, loi=Cytosol_Light) %>%
  pivot_wider(names_from=condition, id_cols=id, values_from=loi) %>%
  filter(is.finite(DMSO), is.finite(Tg)) %>%
  mutate(change=Tg-DMSO) %>%
  merge(RNA_abundance_change, by.y='row.names', by.x='id') %>%
  filter(Tg>0.5) %>%
  ggplot(aes(change, log2FoldChange)) +
  geom_point(alpha=0.5, size=0.5) +
  theme_camprot() +
  xlab('Membrane proportion (Tg/DMSO)') +
  ylab('RNA abundance (Tg/DMSO; log2)')
 

x <- proportions$gene %>%
  merge(RNA_abundance_change, by.y='row.names', by.x='id') %>%
  filter(padj<0.05, log2FoldChange<0) %>%
  arrange(log2FoldChange)

print(x %>% filter(x$id %in% x$id[duplicated(x$id)]))
```



