---
title: "Inspect the nuclear protein profiles"
author:
  - name: "Tom Smith"
    affiliation: "Cambridge Centre for Proteomics"
date: "`r format(Sys.time(), '%d %B, %Y')`"
abstract: | 
  Here, we plot the final marker sets.
output:
  pdf_document:
  html_notebook: default
geometry: margin=1in
fontsize: 11pt
---

```{r}
library(tidyverse)
library(camprotR)
library(pRoloc)
library(pRolocdata)
library(pRolocExt)
library(gplots)
```
Read in the protein quantification with the finalised markers sets
```{r}
combined_protein_res <- readRDS('../../results/combined_protein_res_finalised_markers.rds')
```



```{r, fig.height=4, fig.width=10}


dmso_e <- combined_protein_res$DMSO %>% filterNA() %>% exprs()
labs <- rep(1:8, 3)

colours=get_cat_palette(5)[3:5]

clust <-  dmso_e %>% dist() %>% hclust()

png('../../results/protein_clust.png', width=200, height=80, units='mm', res=500)
heatmap.2(dmso_e,
          dendrogram='row', Colv=FALSE, trace='none', labRow=FALSE, Rowv=clust$order,
          labCol='',
          density.info='none',
          key.title=NA,
          key.xlab='Normalised abundance',
          key.ylab='X',
          xlab=NA,
          breaks=seq(0, 1, length.out=101), 
          col=colorRampPalette(c('grey95', get_cat_palette(2)[2]))(n = 100),
          lhei=c(0.1, 1),
          lwid=c(0.3, 1),
          key=FALSE,
          #keysize=0.05,
          #key.par=list(cex=0.5),
          add.expr = text(x = seq_along(labs), y = -100, cex=1,
                          col=c(rep(colours[1], 8), rep(colours[2], 8), rep(colours[3], 8)),
                          labels = labs, xpd = TRUE),
          key.xtickfun=function() {
                 breaks <- parent.frame()$breaks
                 return(list(
                      at=parent.frame()$scale01(c(breaks[1],
                                                  breaks[51],
                                                  breaks[101])),
                      labels=c(as.character(breaks[1]),
                               as.character(breaks[51]),
                               as.character(breaks[101]))
                      ))
           })

dev.off()
```


First, let's inspect the mean marker set profiles. 
```{r}
for(markers in c('markers', 'markers')){
  
  
  #get order for Control
  .data <- combined_protein_res$DMSO %>%
    filterNA()
  
  .data <- .data %>%
  normalise('sum')
  
  hc <- mrkHClust(.data, plot = FALSE, fcol=markers)
  mm <- getMarkerClasses(.data, fcol=markers)
  ord <- levels(factor(mm))[order.dendrogram(hc)]
  

  for(condition in names(combined_protein_res)){
    
    .data <- combined_protein_res[[condition]]  %>% filterNA()
    
    #colnames(.data) <- 1:10
    hc <- mrkHClust(.data, plot = FALSE, fcol=markers)
    mm <- getMarkerClasses(.data, fcol=markers)
    
    fmat <- mrkConsProfiles(.data, fcol=markers)
    
    p <- plotConsProfiles(fmat, order = ord, plot=FALSE) +
      scale_fill_continuous(name='Sum norm.\nintensity',
                            low = "white", high=get_cat_palette(1),
                            limits=c(0,1),
                            breaks=seq(0,1,0.2)) +
      scale_x_discrete(labels=pData(.data)$fraction) +
      ggtitle(condition) +
      theme_camprot(base_size=15) +
      geom_vline(xintercept=8.5) +
      geom_vline(xintercept=16.5) +
      theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1)) +
      xlab('')
    
    print(p)    
  }
  

}

```

PCAs and linear marker plots
```{r, fig.height=7, fig.width=7}

colours <- getStockcol()

for(condition in names(combined_protein_res)){
  
  if(condition=='Control'){
    position='topleft'
  }
  else{ position = 'bottomleft'
  }
  
  .data <- combined_protein_res[[condition]]

  .data %>% plot2D(col=colours, fcol='markers')
  addLegend(.data, col=colours, where=position, fcol='markers')
  
  .data %>% plot2D(col=colours, dims=c(3,4), fcol='markers')
  addLegend(.data, col=colours, where=position, fcol='markers')
  
  .data %>% plot2D(col=colours, dims=c(2,3), fcol='markers')
  addLegend(.data, col=colours, where=position, fcol='markers')
  
  #.data %>% plot2D(col=colours, method='kpca', fcol='markers')
  #addLegend(.data, col=colours, where=position, fcol='markers')
  
  .data <- .data[fData(.data)$markers!='unknown',]

  alphas <- length(getMarkerClasses(.data, fcol='markers')) * classWeights(.data, fcol='markers')
  
  p <- .data %>%
    plot_marker_profiles(facet_by='markers', group_by='replicate', alpha=alphas, fcol='markers') +
    ggtitle(condition) +
    scale_x_discrete(labels=pData(.data)$fraction) +
    theme_camprot(base_size=12) +
    theme(axis.text.x=element_text(angle=90, vjust=0.5, hjust=1), legend.position='none') +
    geom_vline(xintercept=8.5, linetype=2, lwd=0.25) +
    geom_vline(xintercept=16.5, linetype=2, lwd=0.25) +
    scale_colour_manual(values=colours)

  print(p)
  
}




```

Linear profiles
```{r}
obj <- combined_protein_res$DMSO %>% filterNA()
obj <- obj[fData(obj)$markers!='unknown']

exprs_df <- exprs(obj)
f_df <- fData(obj) %>% dplyr::select(markers)
exprs_df <- exprs_df %>% data.frame()
new_colnames2old_colnames <- data.frame(new = colnames(exprs_df), 
    old = colnames(exprs(obj)))

exprs_df <- exprs_df %>% tibble::rownames_to_column("id") %>% 
    tidyr::pivot_longer(-id, names_to = "sample", values_to = "abundance") %>% 
    merge(new_colnames2old_colnames, by.x = "sample", by.y = "new") %>% 
    mutate(sample = factor(camprotR::remove_x(old), levels = colnames(obj)))
exprs_df <- merge(exprs_df, f_df, by.x = "id", by.y = "row.names") %>% 
    merge(pData(obj), by.x = "sample", by.y = "row.names")

summaried_exprs_df <- exprs_df %>%
  mutate(markers=recode(Hmisc::capitalize(tolower(markers)), 'Er'='ER', 'Pm'='PM')) %>%
  group_by(fraction, markers, replicate) %>%
  summarise('mean'=mean(abundance),  'sd'=sd(abundance))

p <- summaried_exprs_df %>%
  filter(replicate==1) %>%
  ggplot(aes(as.numeric(fraction), colour=markers, fill=markers)) +
  geom_line(aes(y=mean)) +
  geom_ribbon(aes(ymax=mean+(2*sd), ymin=mean-(2*sd)), alpha=0.5, colour=NA) +
  theme_camprot(border=FALSE, base_size=15, base_family='sans') +
  scale_colour_manual(values=colours, guide=FALSE) +
  scale_fill_manual(values=colours, guide=FALSE) +
  scale_x_continuous(breaks=1:8, name='Fraction') +
  ylab('Abundance\n(sum normalised)')

print(p)


p <- summaried_exprs_df %>%
  filter(replicate==1) %>%
  ggplot(aes(as.numeric(fraction), colour=markers, fill=markers)) +
  geom_line(aes(y=mean)) +
  geom_ribbon(aes(ymax=mean+2*sd, ymin=mean-2*sd), alpha=0.5, colour=NA) +
  facet_wrap(~markers, ncol=2) +
  theme_camprot(border=FALSE, base_size=10, base_family='sans') +
  scale_fill_manual(values=colours, guide=FALSE) +
  scale_colour_manual(values=colours, name='') +
  theme(strip.text=element_blank(),
        strip.background=element_blank()) +
  scale_x_continuous(breaks=1:8, name='Fraction') +
  scale_y_continuous(breaks=seq(0,1,0.5), name='Abundance (sum normalised)')

print(p)
ggsave('../../../../5_manuscript_figures/Figure_1/protein_profiles/linear.png', height=4, width=4)
ggsave('../../../../5_manuscript_figures/Figure_1/protein_profiles/linear.pdf', height=4, width=4)


p <- summaried_exprs_df %>%
  ggplot(aes(as.numeric(fraction), colour=markers, fill=markers)) +
  geom_line(aes(y=mean)) +
  facet_wrap(~replicate, nrow=1) +
  theme_camprot(border=FALSE, base_size=15, base_family='sans') +
  scale_colour_manual(values=colours, name='') +
  theme(strip.text=element_blank(),
        strip.background=element_blank()) +
  scale_x_continuous(breaks=1:8, name='Fraction') +
  ylab('Abundance\n(sum normalised)')

print(p)

ggsave('../../../../5_manuscript_figures/Figure_1/protein_profiles/linear_all.png', height=3, width=7)
ggsave('../../../../5_manuscript_figures/Figure_1/protein_profiles/linear_all.pdf', height=3, width=7)

```


```{r}
colours <- readRDS('../../../../6_shiny_app/out/shiny_colours.rds')$Protein %>% unname()

for(condition in names(combined_protein_res)){
  
  if(condition=='Control'){
    position='topleft'
  }
  else{ position = 'bottomleft'
  }
  
  .data <- combined_protein_res[[condition]]

  .data <- .data[fData(.data)$markers!='unknown',]

  p <- .data %>%
    plot_marker_profiles(group_by='replicate', plot_all=FALSE, fcol='markers') +
    ggtitle(condition) +
    scale_x_discrete(labels=pData(.data)$fraction) +
    theme_camprot(base_size=15) +
    #theme(axis.text.x=element_text(angle=90, vjust=0.5, hjust=1), legend.position='none') +
    geom_vline(xintercept=8.5, linetype=2, lwd=0.25) +
    geom_vline(xintercept=16.5, linetype=2, lwd=0.25) +
    scale_colour_manual(values=colours)

  print(p)
}

```
OK, too busy for this sort of representation to work. Let's just use the t-SNE (see bottom of notebook)





Kernal PCAs
```{r, fig.height=6, fig.width=6}
for(condition in names(combined_protein_res)){
  pdf(sprintf('../../../../5_manuscript_figures/Figure_1/protein_profiles/kernal_pca_%s.pdf', condition),
      height=5, width=5)
  combined_protein_res[[condition]] %>% plot2D(main = paste(condition, ' - ', "Kernal PCA"),
                                               grid=FALSE, method='kpca', col=colours, fcol='markers')
  dev.off()
}
```


```{r, fig.height=10, fig.width=10, eval=FALSE}

perps <- c(5, 10, 25, 50)

for(condition in names(combined_protein_res)){
  
    par(mfrow = c(3, 2))
    combined_protein_res[[condition]] %>% plot2D(main = paste(condition, ' - ', "PCA"), col=colours, fcol='markers')
    
    sapply(perps,
           function(perp) {
             combined_protein_res[[condition]] %>% normalise('sum') %>%
               plot2D(method = "t-SNE", methargs = list(perplexity = perp), col=colours, fcol='markers')
             title(main = paste(condition, ' - ', "t-SNE, perplexity", perp))
           })
}

```


To avoid having to recompute t-SNE in the future, we will select a singe perplexity and add the projections to the fData. We'll use perplexity==25 for both conditions
```{r}
set.seed(0)
perp <- 25


combined_protein_res_with_tsne <- combined_protein_res %>% lapply(function(x){
  
  tsne_proj <- make_proj(x, method='t-SNE', methargs = list(perplexity = perp))
  
  fData(x)[,c('Dimension 1', 'Dimension 2')] <- tsne_proj[
    rownames(fData(x)),c('Dimension 1', 'Dimension 2')]
  
  return(x)

})
```


```{r}
combined_protein_res_with_tsne %>% names() %>% sapply(function(condition){
  p <-combined_protein_res_with_tsne[[condition]] %>%
    fData() %>%
    data.frame() %>%
    arrange(markers!='unknown') %>%
    mutate(markers=Hmisc::capitalize(tolower(markers))) %>%
    mutate(markers=recode(markers, 'Pm'='PM', 'Er'='ER')) %>%
    ggplot(aes(Dimension.1, Dimension.2)) +
    xlab('Dimension 1') +
    ylab('Dimension 2') +
    theme_camprot(border=FALSE, base_size=15, base_family='sans') +
    guides(alpha=FALSE, size=FALSE) +
    geom_point() +
    aes(colour=markers, alpha=markers=='Unknown', size=markers=='Unknown') +
    scale_alpha_manual(values=c(1, 0.25)) +
    scale_size_manual(values=c(1, 0.5)) +
    scale_colour_manual(name='', values=c(colours[1:12], 'grey80'))
  
  print(p)
  
  ggsave(sprintf('../../../../5_manuscript_figures/Figure_1/protein_profiles/%s_tsne.png', condition))
  ggsave(sprintf('../../../../5_manuscript_figures/Figure_1/protein_profiles/%s_tsne.pdf', condition))

})
```

```{r}
saveRDS(combined_protein_res_with_tsne, '../../results/combined_protein_res_with_tsne.rds')
```



