---
title: "Isoform localisation"
author:
  - name: "Tom Smith"
    affiliation: "Cambridge Centre for Proteomics"
date: "`r format(Sys.time(), '%d %B, %Y')`"
abstract: | 
  Here,we identify examples of transcript isoforms with differential localisation or differential relocalisation
output:
  pdf_document:
  html_notebook: default
geometry: margin=1in
fontsize: 11pt
---

```{r}
library(tidyverse)
library(camprotR)
library(biomaRt)
library(pRoloc)
library(ggtern) # note this means plotting non ternary ggplots using using ggplot2:::plot.ggplot(p)
```

Read in proportions
```{r}
proportions_gene <- readRDS('../4_proportional_localisation/results/proportions.rds')$gene
proportions <- readRDS('../4_proportional_localisation/results/proportions.rds')$transcript

proportions_dmso <- proportions %>% filter(condition=='DMSO')

proportions_tg <- proportions %>% filter(condition=='Tg')
```

```{r}
proportions_all_reps <- readRDS('../4_proportional_localisation/results/proportions_per_rep.rds')$transcript
```

```{r}
ensembl <- biomaRt::useEnsembl('ensembl', version=102, dataset='hsapiens_gene_ensembl')
x <-listAttributes(ensembl)
x %>% filter(grepl('seq', description, ignore.case=TRUE))
transcript2biotype <- getBM(attributes=c('transcript_biotype', 'ensembl_transcript_id', 'ensembl_gene_id', 'external_transcript_name'), mart=ensembl)


head(transcript2biotype)
```

Subset to tx from genes with multiple isoforms
```{r}
proportions_dmso_multiple_isoforms <- proportions_dmso %>%
  merge(transcript2biotype, by.x='id', by.y='ensembl_transcript_id') %>%
  filter(transcript_biotype=='protein_coding') %>%
  filter((duplicated(ensembl_gene_id)|duplicated(ensembl_gene_id, fromLast=TRUE)))

proportions_tg_multiple_isoforms <- proportions_tg %>%
  merge(transcript2biotype, by.x='id', by.y='ensembl_transcript_id') %>%
  filter(transcript_biotype=='protein_coding') %>%
  filter((duplicated(ensembl_gene_id)|duplicated(ensembl_gene_id, fromLast=TRUE)))

```

Find the genes with the biggest difference in granule proportions in DMSO
```{r}
isoform_specific_granule_loc <- proportions_dmso %>%
  merge(transcript2biotype, by.x='id', by.y='ensembl_transcript_id') %>%
  filter(transcript_biotype=='protein_coding', sd_Granule<0.1) %>%
  filter((duplicated(ensembl_gene_id)|duplicated(ensembl_gene_id, fromLast=TRUE))) %>%
  group_by(ensembl_gene_id) %>%
  summarise(range_gra=max(Granule)-min(Granule)) %>%
  arrange(desc(range_gra))
```

Find the genes with the biggest difference in nucleus proportions in DMSO
```{r}
isoform_specific_nuc_loc <- proportions_dmso_multiple_isoforms %>%
  mutate(Nuc=Nucleus+Nucleolus) %>%
  group_by(ensembl_gene_id) %>%
  summarise(range_nuc=max(Nuc)-min(Nuc)) %>%
  arrange(desc(range_nuc))
```

Find the genes with the biggest difference in membrane proportions in DMSO
```{r}
isoform_specific_mem_loc <- proportions_dmso_multiple_isoforms %>%
  group_by(ensembl_gene_id) %>%
  summarise(range_mem=max(Membrane)-min(Membrane)) %>%
  arrange(desc(range_mem))
```

Read in the quant data
```{r}
lorna_obj <- readRDS('../4_proportional_localisation/results/all_datasets_final.rds')

lorna_dmso_tx <- lorna_obj$transcript$DMSO
lorna_tg_tx <- lorna_obj$transcript$Tg
#readRDS('../3_clustering/out/tg_with_tsne_tx.rds')
```



```{r}
library(ensembldb)
library(AnnotationHub)
library(Gviz)

# use line below to reforce hub download
#refreshHub(hubClass="AnnotationHub")
hub <- AnnotationHub()

# 102 is not in this version
query(hub, c("EnsDb", "sapiens", "101"))
ensdb <- hub[['AH83216']]

filter <- dplyr::filter
```

Function for transcript tracks
```{r}
#tx_foi must be a named list. Names=ensembl_transcript_ids, values=transcript_names
plot_tx_models <- function(tx_foi, just_gene_models=FALSE){
  
  txs <- getGeneRegionTrackForGviz(ensdb, filter = ~ tx_id %in% names(tx_foi))
  
  contig <- paste0('chr', as.character(txs@seqnames)[[1]])
  
  ideo_track <- IdeogramTrack(genome = "hg38", chromosome = contig)
  ## - Genome axis
  gaxis_track <- GenomeAxisTrack()
  ## - Transcripts
  txs@elementMetadata$symbol <- tx_foi[txs@elementMetadata$transcript]
  gene_track <- GeneRegionTrack(txs,
                                gene=txs@elementMetadata$transcript,
                                showId = TRUE,
                                #just.group = "right",
                                geneSymbol = FALSE,
                                fontsize = 20,
                                name='Tx models')
  ?GeneRegionTrack
  ## Generate the plot
  if(just_gene_models){
    p <- plotTracks(list(gene_track))
  } else{
    p <- plotTracks(list(ideo_track, gaxis_track, gene_track))
  }
  
  return(p)
}
```




Function to plot PCA for transcripts
```{r}
plot_tx_pca <- function(tx_foi){
  .pca <- plot2D(lorna_obj$transcript$DMSO$corrected, addLegend='topright')
  highlightOnPlot(.pca, tx_foi)
  
  .pca <- plot2D(lorna_obj$transcript$Tg$corrected, addLegend='topleft')
  highlightOnPlot(.pca, tx_foi)
}
```


Plots for top genes with clearest difference between transcripts
```{r}
for(gene_id in head(isoform_specific_nuc_loc$ensembl_gene_id, 20)){
  
  tx <- proportions_dmso_multiple_isoforms %>% filter(ensembl_gene_id==gene_id)
  
  tx_foi <- tx$external_transcript_name
  names(tx_foi) <- tx$id
  
  plot_tx_models(tx_foi)
  
  .data_gene <- proportions_gene %>%
    filter(id == tx$ensembl_gene_id[[1]]) %>%
    filter(condition=='DMSO') %>%
    mutate(Nuc=Nucleolus+Nucleus,
           Cyt=Cytosol,
           Gra=Granule)
  
  .data <- proportions_dmso_multiple_isoforms %>%
    filter(id %in% names(tx_foi)) %>%
    mutate(Nuc=Nucleolus+Nucleus,
           Cyt=Cytosol,
           Gra=Granule)

  p <- .data %>%
    ggtern(aes(x=Cyt+Gra, y=Membrane, z=Nuc)) +
    theme_bw(base_size=10) +
    xlab('Cyt.') +
    ylab('Mem.') +
    zlab('Nuc.') +
    theme_arrowlarge() +
    theme_nogrid() +
    theme_hidetitles() +
    theme(strip.background = element_blank(),
          strip.text.x=element_text(size=15, vjust=2),
          strip.text.y=element_text(size=15, vjust=2)) + 
    geom_point(aes(colour=external_transcript_name), size=4)  +
    geom_point(data=.data_gene, size=4, pch=8)  +
    scale_colour_manual(values=get_cat_palette(length(tx_foi)), name='')
  
  print(p)
}
```

```{r}
for(gene_id in head(isoform_specific_mem_loc$ensembl_gene_id, 5)){
  
  tx <- proportions_dmso_multiple_isoforms %>% filter(ensembl_gene_id==gene_id)
  
  tx_foi <- tx$external_transcript_name
  names(tx_foi) <- tx$id
  
  plot_tx(tx_foi)
  
  .data_gene <- proportions_gene %>%
    filter(id == tx$ensembl_gene_id[[1]]) %>%
    filter(condition=='DMSO') %>%
    mutate(Nuc=Nucleolus+Nucleus,
           Cyt=Cytosol,
           Gra=Granule)
  
  .data <- proportions_dmso_multiple_isoforms %>%
    filter(id %in% names(tx_foi)) %>%
    mutate(Nuc=Nucleolus+Nucleus,
           Cyt=Cytosol,
           Gra=Granule)

  p <- .data %>%
    ggtern(aes(x=Cyt+Gra, y=Membrane, z=Nuc)) +
    theme_bw(base_size=15) +
    xlab('Cyt.') +
    ylab('Mem.') +
    zlab('Nuc.') +
    theme_arrowlarge() +
    theme_nogrid() +
    theme_hidetitles() +
    theme(strip.background = element_blank(),
          strip.text.x=element_text(size=15, vjust=2),
          strip.text.y=element_text(size=15, vjust=2)) +
    geom_point(aes(colour=external_transcript_name), size=6)  +
    geom_point(data=.data_gene, size=6, pch=8)
  
  if(length(tx_foi)<=13){
    p <- p + 
      scale_colour_manual(values=get_cat_palette(length(tx_foi)), name='')
  }
  
  return(p)
}
```
OK, so the last gene above is really interesting. Very little difference in the coding sequence. However, 201 has an insertion of 52-52: Q → HPLPLCR and 208 is truncated and has an alternative C-terminus (PPFPHDPLLSGLPSAT). Given the canonical polypeptide is relatively short (140 AA), it's not suprising it doesn't reach the ER during translation. However, the non-canonical isoforms do! As both the alternative sequences are relatively proline-rich, I wonder if ribosome stalling is the explanation?

```{r}
transcript2coding <- getBM(attributes=c('ensembl_transcript_id', 'coding', 'external_transcript_name'), mart=ensembl,
                           filters = 'ensembl_transcript_id', values=proportions_dmso_multiple_isoforms$id)


dim(transcript2coding)
transcript2coding <- transcript2coding %>% filter(coding!="Sequence unavailable")
dim(transcript2coding)
transcript2coding <- transcript2coding %>% filter(!grepl('N', coding))
dim(transcript2coding)

coding_seq <- transcript2coding$coding %>%
  gsub(pattern='T', replacement='U') %>%
  Biostrings::RNAStringSet()

polypep <- suppressWarnings(coding_seq %>% Biostrings::translate())
length(polypep)

aa_freq <- Biostrings::letterFrequency(polypep, letters=Biostrings::alphabet(polypep)) %>%
  data.frame()

aa_freq$aas <- rowSums(aa_freq)
dip_freq <- Biostrings::vcountPattern('PP', polypep)

rownames(aa_freq) <- transcript2coding$ensembl_transcript_id
names(dip_freq) <- transcript2coding$ensembl_transcript_id
```

```{r}
tx <- proportions_dmso_multiple_isoforms %>% filter(ensembl_gene_id==gene_id)
  
tx_foi <- tx$external_transcript_name
names(tx_foi) <- tx$id

plot_tx(tx_foi)
aa_freq[names(tx_foi),'P']
dip_freq[names(tx_foi)]

```

```{r}
tx <- proportions_dmso_multiple_isoforms %>% filter(ensembl_gene_id==gene_id)
  
tx_foi <- tx$external_transcript_name
names(tx_foi) <- tx$id

plot_tx_models(tx_foi, just_gene_models=TRUE)

png('../../../../../5_manuscript_figures/Figure_2/transcripts/example_gene_models.png', width=4, height=4, units='in', res=400)
plot_tx_models(tx_foi, just_gene_models=TRUE)
dev.off()


pdf('../../../../../5_manuscript_figures/Figure_2/transcripts/example_gene_models.pdf', width=4, height=4)
plot_tx_models(tx_foi, just_gene_models=TRUE)
dev.off()
```
Ternary plots for TMEM
```{r}
.data_gene <- proportions_gene %>%
    filter(id == tx$ensembl_gene_id[[1]]) %>%
    filter(condition=='DMSO') %>%
    mutate(Nuc=Nucleolus+Nucleus,
           Cyt=Cytosol,
           Gra=Granule)
  
.data <- proportions_dmso_multiple_isoforms %>%
  filter(id %in% names(tx_foi)) %>%
  mutate(Nuc=Nucleolus+Nucleus,
         Cyt=Cytosol,
         Gra=Granule)

p <- .data %>%
  ggtern(aes(x=Cyt+Gra, y=Membrane, z=Nuc)) +
  theme_bw(base_size=15) +
  xlab('Cyt.') +
  ylab('Mem.') +
  zlab('Nuc.') +
  theme_arrowlarge() +
  theme_nogrid() +
  theme_hidetitles() +
  theme(strip.background = element_blank(),
        strip.text.x=element_text(size=15, vjust=2),
        strip.text.y=element_text(size=15, vjust=2)) +
  geom_point(aes(colour=external_transcript_name), size=10)  +
  geom_point(data=.data_gene, size=10, pch=8) + 
  scale_colour_manual(values=get_cat_palette(length(tx_foi)), name='')

print(p)
ggsave('../../../../../5_manuscript_figures/Figure_2/transcripts/example_tern.png', width=6, height=3.5)
ggsave('../../../../../5_manuscript_figures/Figure_2/transcripts/example_tern.pdf', width=6, height=3.5)
```
